---
title: "JIBE Melbourne: COPD incidence data"
author: "Chris Jackson Belen Zapata-Diomedi, Steve Pemberton"
email: "chris.jackson@mrc-bsu.cam.ac.uk"
date: today
format: 
  html:
    code-fold: true
    embed-resources: true
    df-print: paged
execute: 
    message: false
    warning: false
toc: true
number-sections: true
editor: visual
---

This script produces estimates of chronic obstructive pulmonary disease (COPD) incidence, by cby age, sex and area socioeconomic quintile, for Australia using data for 2016 and 2018.

The data is obtained from the Global Burden of Disease for disease incidence by age and sex and from AIHW by sex and socioeconomic status ([JIBE data summary spreadsheet](https://gbr01.safelinks.protection.outlook.com/ap/x-59584e83/?url=https%3A%2F%2Frmiteduau.sharepoint.com%2F%3Ax%3A%2Fr%2Fsites%2FJIBEUKAUS%2FShared%2520Documents%2FWP%25206%2Fhealth%2Fdata%2Fdiseases%2520data.xlsx%3Fd%3Dwed9fbef6faa046df8f4e6453f1a1463b%26csf%3D1%26web%3D1%26e%3DhKUlh4&data=05%7C02%7Cchris.jackson%40mrc-bsu.cam.ac.uk%7Cdfd4effdc1c346447c0508dc30f7a9ae%7C513def5bdf174107b5523dba009e5990%7C0%7C0%7C638439087576234119%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=WdIPqRrgjJze3vSbAhvWQYZO3K1R2IUlDXcxbWXFVJA%3D&reserved=0)). There is no unique source of data for what we need, hence the combination of data sources. Source files:

-   Incidence by 5-year age group and sex for COPD:

    -   `gbd_incidence.csv` (available from GBD data tool, data is for 2018, baseline year for the jibe model)

-   Prevalence by socioeconomic area for 2017-18 (we do not have data for incidence by socioeconomic area, hence we use prevalence):

    -   `copd-prevalence.xlsx`, which contains Prevalence of chronic obstructive pulmonary disease among people aged 45 and over, by sex, remoteness and socioeconomic area, 2017–18: Table 1.2
    
These data refer to the whole of Australia, and we do not have area specific data (e.g. SA2 as we do with mortality). These data are combined to produce estimates for COPD by age, sex and area socioeconomic quintile. This assumes that the effect of socioeconomic status on COPD is the same for all age and sex groups.


```{r}
library(tidyverse)
library(readxl)
```

```{r setup, include=FALSE}
# knitr::opts_knit$set(root.dir = "C:/Users/beluz/OneDrive - RMIT University/JIBE/JIBE-WP6/healthmicrosim/")
knitr::opts_knit$set(root.dir = "C:/Users/steve/Documents/MEGAsync/Steve/RMIT JIBE/health_microsim/")
knitr::opts_chunk$set(cache = TRUE)
```

# Incidence by age and sex

Data by year of age are assumed to be the same within the published five-year age groups.

```{r}
gbd <- read.csv("health/original/GBD/gbd_incidence.csv") %>%
  filter(measure %in% "Incidence") %>%
  filter(cause %in% "Chronic obstructive pulmonary disease") %>%
  filter(metric %in% "Rate") %>%
  filter(year %in% 2018) %>%
  mutate(age = gsub(" years", "", age)) %>%
  mutate(age = case_when(age == "<5"  ~ "0-4",
                         age == "95+" ~ "95-99",
                         .default = age)) %>%
  separate_wider_delim(age, "-", names = c("startage","stopage")) %>%
  mutate(across(startage:stopage, as.numeric),
         agerange = stopage - startage + 1) %>%
  arrange(startage) %>%
  group_by(sex) %>%
  slice(rep(1:n(), agerange)) %>%
  mutate(age = row_number() - 1) %>%
  ungroup()
stopifnot(gbd$age >= gbd$startage, gbd$age <= gbd$stopage)
copdinc_agesex <- gbd %>%
    select(cause, age, sex, rate = val)
```


# Incidence by socioeconomic status and sex

-   Socioeconomic areas are classified according to population-based quintiles using the Index of Relative Socio-Economic Disadvantage (IRSD) based on Statistical Area Level 2 (SA2) of usual residence
-   As noted above, the data is for prevalence rather than incidence, as incidence data by socioeconomic status is not available. 
-   This prevalence data is used to adjust GBD incidence data by age and sex, allowing it to be further dissagregated by socioeconomic status.

```{r}
copd_socio <- read_xlsx("health/original/AIHW/copd-prevalence.xlsx", sheet="Table 1.2", skip=2, 
                        col_names = TRUE, n_max=10) %>%
  slice(-c(1:5)) %>%
  rename_with( ~ c("socio", "Male", "MaleCI", "Female", "FemaleCI", "Person", "PersonCI")) %>%
  dplyr::select(socio, Male, Female) %>%
  pivot_longer(cols = c("Male", "Female"), names_to = "sex", values_to = "rate") %>%
  mutate(socio = case_when(socio == "1 (lowest)" ~ "1", # 1: lowest, to 5: highest
                           socio == "5 (highest)" ~ "5",
                           .default = socio),
         socio = as.numeric(socio)) %>%
  group_by(sex) %>%
  mutate(rr = rate / first(rate)) %>%
  ungroup() %>%
  select(socio, sex, rate, rr)
```

# Estimating incidence by combinations of risk factors

As in the Melbourne mortality data analysis (section 2.2) to disaggregate an event rate (incidence by age and sex in this case) by a further risk factor, given

-   the rate $r_{ave}$ by age and sex, averaged over people with or without the risk factor

-   a relative (unstandardised) rate $RR$ between people with/without the risk factor

-   the proportions of people with the risk factor $p_1$ and without the risk factor $p_0$

we can estimate the rates with ($r_1$) and without the risk factor ($r_0$), because if $r_{ave} = p_1 r_1 + p_0 r_0 = p_1 r_0 RR + p_0 r_0$,we can estimate $r_0 = r_{ave} / (p_1 RR + p_0)$

This generalises to categorical risk factors with more than two categories, in this case, the area socioeconomic indicator. Given

$r_{ave} = r_1 \sum_{i=1}^G p_i RR_i$

we can compute $r_1$ given $r_{ave}$, the socioeconomic relative rates $RR_i$ and socioeconomic population proportions $p_i$. Hence we can compute the remaining $r_i$ by multiplying $r_1$ by $RR_i$.

## Implementation of disaggregating risks for cancer incidence

We assume that $p_i = 1/5$ is the proportion of people in each of the five quintiles of socioeconomic status, and these proportions are the same for each age/sex group. The quintiles are defined so that the same number of areas falls in each quintile. So while this assumption will not be exact, it is expected to be approximately true. To get better estimates of these proportions, we would need data on the population by age, sex and SES quintile.

The code to perform this disaggregation works by stretching out the age/sex incidence data vertically to create a dataset with one row per combination of age, sex and socioeconomic quintile. This is then joined with the dataset `copd_socio` on relative incidence by socioeconomic quintiles. The term $\sum_{i=1}^G p_i RR_i$ is then computed as `sum_pRR`, and joined to the streched dataset `copdinc`.  This allows the disaggregated rates `rate` to be computed in a vectorised way.




```{r}
copdinc <- copdinc_agesex %>% 
    slice(rep(row_number(), each=5)) %>%
    mutate(socio = rep(1:5, length.out=n())) %>%
    left_join(copd_socio %>% select(sex, socio, rr_socio = rr), 
              by = c("sex", "socio")) %>%
    mutate(p_socio = 1/5,    # assumption
           pRR = p_socio * rr_socio,
           socio = ordered(socio))

sum_p <- copdinc %>% 
    group_by(age, sex) %>%
    summarise(sum_pRR = sum(pRR), .groups="drop")
    
copdinc <- copdinc %>%
    left_join(sum_p, by=c("age", "sex")) %>%
    mutate(rate_soc = rate * rr_socio / sum_pRR, 
           socio = ordered(socio, labels = c("1 (most deprived)",
                                             "2","3","4",
                                             "5 (least deprived)")),
           sex = factor(sex)) 
saveRDS(copdinc, file="cvd_dia_copd_incidence_test.rds")
```
We are left with the incidence by age, sex and socioeconomic status in the variable `rate`. The final dataset is saved in the file `cancer_incidence.rds`, and the first few rows are shown here.

```{r}
knitr::kable(head(copdinc))
```

# Illustration of incidence estimates

Incidence by age, sex and socioeconomic quintile for COPD.

The published incidence by age and sex (averaged over deprivation levels) is shown by the dotted black lines. The estimated incidence disaggregated by socioeconomic quintile is shown by the coloured lines, with socioeconomic status indicated by the lightness of the lines.

This shows that the relative risk between socioeconomic quintiles is assumed to be the same between different age/sex groups, but the average of the deprivation-specific estimates is constrained to be the same as the published average.

Note the very low incidence of cases below approximately age 40, the higher incidence for males than females, and the strong influence of socio economic effect, particularly the much higher rates at the highest level of deprivation.

```{r}
#| fig-height: 12
ggplot(copdinc, 
       aes(x=age, y=rate_soc, col=sex, alpha=socio)) + 
    geom_line(lwd=1.5) + 
    geom_line(data=copdinc_agesex,
              aes(x=age, y=rate, group=sex), 
              inherit.aes = FALSE,
              lwd=1, lty=2, alpha=0.4, col="black") +
    # facet_wrap(~site, ncol=2, scales="free_y") + 
    ylab("Incidence rate (per 100,000)") +
    xlab("Age (years)") +
    coord_cartesian(xlim=c(25, 90)) +
    guides(alpha=guide_legend(title=NULL),
           col=guide_legend(title=NULL)) +
    theme(legend.position = "bottom")
```

