---
title: "JIBE Melbourne: Cancer incidence data"
author: "Chris Jackson"
email: "chris.jackson@mrc-bsu.cam.ac.uk"
date: today
format: 
  html:
    code-fold: true
    embed-resources: true
    df-print: paged
execute: 
    message: false
    warning: false
toc: true
number-sections: true
editor: visual
---

This script produces estimates of cancer incidence, by cancer site, age, sex and area socioeconomic quintile, for Australia in 2016.

The data is obtained from AIHW, as linked from [JIBE data summary spreadsheet](https://gbr01.safelinks.protection.outlook.com/ap/x-59584e83/?url=https%3A%2F%2Frmiteduau.sharepoint.com%2F%3Ax%3A%2Fr%2Fsites%2FJIBEUKAUS%2FShared%2520Documents%2FWP%25206%2Fhealth%2Fdata%2Fdiseases%2520data.xlsx%3Fd%3Dwed9fbef6faa046df8f4e6453f1a1463b%26csf%3D1%26web%3D1%26e%3DhKUlh4&data=05%7C02%7Cchris.jackson%40mrc-bsu.cam.ac.uk%7Cdfd4effdc1c346447c0508dc30f7a9ae%7C513def5bdf174107b5523dba009e5990%7C0%7C0%7C638439087576234119%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=WdIPqRrgjJze3vSbAhvWQYZO3K1R2IUlDXcxbWXFVJA%3D&reserved=0). Source files:

-   `cancer-incidence.xlsx` (published Dec 2021): By cancer site, 5-year age group and sex.

-   `cancer_socioeconomic.xlsx` (published Dec 2021): Table S10.7. By site and area socioeconomic quintile.

These data refer to the whole of Australia. GBD does not provide any further geographical detail, and is likely to have drawn their data on Australia from the same source, so GBD is not considered further here.

These data are combined to produce estimates by cancer site, age, sex and area socioeconomic quintile. This assumes that the effect of socioeconomic status on cancer incidence is the same for all age and sex groups, though this effect differs by cancer site.

```{r}
library(tidyverse)
library(readxl)
```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/beluz/OneDrive - RMIT University/JIBE/JIBE-WP6/healthmicrosim/")

```

# Harmonise definitions of cancer sites

The cancer sites are reported differently between the AIHW tables on incidence by age/sex and incidence by socioeconomic quintiles.

This table details the assumptions that are made in order to match the corresponding rates between the tables. The sites that we intend to use are defined by `stdname` and `ICD10`. Each row of the table below details the site from the AIHW tables that these are assumed to be matched to.

```{r}

sites <- rbind( 
    ## Standard name,   Name in age/sex data,  Name in socio data,   Our intended ICD10,
    c("Breast cancer",   "Breast cancer"  ,      "Breast cancer in females (C50)" , "C50"),
     c("Endometrial cancer", "Endometrial cancer", "Uterine cancer (C54–C55)" ,  "C54–C55"),
      c("Lung cancer",       "Lung cancer",         "Lung cancer (C33–C34)",  "C33–C34"),
      c("Colon cancer",     "Colon cancer",         "Colorectal cancer (C18–C20)", "C18"),
      c("Esophageal cancer", "Oesophageal cancer" , NA,  "C15"),  
      c("Liver cancer",      "Liver cancer",       "Liver cancer (C22)", "C22"),
      c("Kidney cancer",     "Kidney cancer",    "Kidney cancer (C64)", "C64"),
      c("Gastric cardia cancer",  "Stomach cancer", NA, "C16"),
      c("Myeloid leukemia", "Chronic myeloid leukaemia (CML)"  , NA, "C92.1"),
      c("Myeloma", "Multiple myeloma", NA, "C90.0"),
      c("Head and neck cancer", "Head and neck cancer (including lip)", "Head and neck cancers (with lip) (C00–C14, C30–C32)", "C01–C14, C30–C32"),
      c("Rectum cancer", "Rectal cancer (including rectosigmoid junction)", "Colorectal cancer (C18–C20)", "C20"),
      c("Bladder cancer", "Bladder cancer", "Bladder cancer (C67)", "C67"),
      c("Prostate cancer", "Prostate cancer", "Prostate cancer (C61)", "C61"),
      c("Malignant melanoma", "Melanoma of the skin", "Melanoma of the skin (C43)", "C43"),
    c("All cancers", "All cancers combined", "All cancers combined (C00–C97, D45, D46, D47.1, D47.3–D47.5)", "C00–C97, D45, D46, D47.1, D47.3–D47.5")
    ) |> 
    as_tibble() |>
    setNames(c("stdname","agesexname","socioname","ICD10") ) |>
    select(stdname, ICD10, agesexname, socioname)
knitr::kable(sites, col.names = c("Standardised name","ICD 10 code",
             "Closest site in age/sex data","Closest site in socioeconomic data"))

# assume unique row in age-sex data for each standard site
stopifnot(!any(duplicated(sites$agesexname))) 
```

# Incidence by age and sex

Data by year of age are assumed to be the same within the published five-year age groups.

```{r}
## #| cache: true

inc <- read_xlsx("health/original/AIHW/cancer-incidence.xlsx", sheet="Table S1a.1", skip=5, col_names=TRUE) |> 
    filter(Year==2016) |>
    select(sex = Sex, age = "Age group (years)", 
           site = `Cancer group/site`, 
           rate = "Age-specific rate\r\n(per 100,000)") |>
    filter(age != "All ages combined",
           sex != "Persons") |>
    mutate(age = ifelse(age=="90+","90–95",age)) |>
    separate_wider_delim(age, "–", names = c("startage","stopage")) |>
      mutate(across(startage:stopage, as.numeric),
             agerange = stopage - startage + 1) |>
    group_by(sex, site) |>
    slice(rep(1:n(), agerange)) |>
    mutate(age = row_number() - 1,
           site = sites$stdname[match(site, sites$agesexname)]) |>
    filter(site %in% sites$stdname)
stopifnot(inc$age >= inc$startage, inc$age <= inc$stopage)
cancerinc_agesex <- inc |> 
    select(site, age, sex, rate)
```

# Incidence by socioeconomic status

-   Presumably (though not indicated in the source) socioeconomic status describes 2016 SA2 areas.

-   As per Cancer 2021 report socioeconomic status corresponds to the 2016 area SA2 of usual residence (Appendinx G of <https://www.aihw.gov.au/getmedia/0ea708eb-dd6e-4499-9080-1cc7b5990e64/aihw-can-144.pdf?v=20230605165731&inline=true>). (BZD 07/03/2024)

-   For the cancer sites with missing socioeconomic gradients (esophageal, gastric, myeloid leukemia, myeloma), we will assume that socioeconomic gradient is the same as that for all cancers combined. This may not be true however. The socioeconomic gradient for all cancers is not large - this is largely because the most common cancers (breast, lung) have opposite gradients that cancel out.

-   Socioeconomic gradients are published for all colorectal cancer combined. These are assumed to be the same for colon and rectal cancer.

```{r}
## #| cache: true

# cancerinc_socio <- read_xlsx(paste0(data_dir,"cancer_socioeconomic.xlsx"), sheet="Table S10.7", 
#                              skip=6, col_names = TRUE, n_max=90) |>
#     rename(site = `Cancer site/type (ICD-10 codes)`) |>
#     fill(site) |>
#     select(site, socio = "Socioeconomic area",
#            rate = "ASR (per 100,000)...5") |>
#     filter(socio != "Australia") |>
#     mutate(socio = case_when(socio=="1 Lowest" ~ "1", # 1: lowest, to 5: highest
#                              socio=="5 Highest" ~ "5",
#                              .default = socio),
#            socio = as.numeric(socio), 
#            rr = rate / rate[1], # BZD: all rates are dividing by the first rate, I shoulde be by diseases, I modified. 
#            site = sites$stdname[match(site, sites$socioname)]) |>
#     filter(site %in% sites$stdname)

## Above code modified to calculate RR in reference to each disease (BZD-07/03/2024)
## When comparing with original calclations from Chris from file cancer_incidence seems that this change makes no difference in final results. 
cancerinc_socio <- read_xlsx("health/original/AIHW/cancer_socioeconomic.xlsx", sheet="Table S10.7", 
                             skip=6, col_names = TRUE, n_max=90) |>
    rename(site = `Cancer site/type (ICD-10 codes)`) |>
    fill(site) |>
    select(site, socio = "Socioeconomic area",
           rate = "ASR (per 100,000)...5") |>
    filter(socio != "Australia") |>
    mutate(socio = case_when(socio=="1 Lowest" ~ "1", # 1: lowest, to 5: highest
                             socio=="5 Highest" ~ "5",
                             .default = socio),
           socio = as.numeric(socio)) |>
  group_by(site) |>
  mutate(rr = rate / first(rate)) |>
  ungroup() |>
  mutate(site = sites$stdname[match(site, sites$socioname)])  |>
    filter(site %in% sites$stdname)

####

sociona_sites <- c("Esophageal cancer","Gastric cardia cancer","Myeloid leukemia","Myeloma")
nadf <- cancerinc_socio |> 
    filter(site == "All cancers") |>
    slice(rep(1:n(), length(sociona_sites))) |>
    mutate(site = rep(sociona_sites, each=5))

## For rectum cancer, assume same socioeconomic distribution as all colorectal cancer
dup_sites <- sites$stdname[!is.na(sites$socioname) & duplicated(sites$socioname)]
## Assume this is only one with a coarser site definition in the socioeconomic data
stopifnot(dup_sites == "Rectum cancer") # handle specially

dupdf <- cancerinc_socio |> 
    filter(site == "Colon cancer") |>
    mutate(site = "Rectum cancer")

cancerinc_socio <- cancerinc_socio |> 
    rbind(nadf) |>
    rbind(dupdf)



```

# Estimating incidence by combinations of risk factors

As in the Melbourne mortality data analysis (section 2.2) to disaggregate an event rate (incidence by age and sex in this case) by a further risk factor, given

-   the rate $r_{ave}$ by age and sex, averaged over people with or without the risk factor

-   a relative (unstandardised) rate $RR$ between people with/without the risk factor

-   the proportions of people with the risk factor $p_1$ and without the risk factor $p_0$

we can estimate the rates with ($r_1$) and without the risk factor ($r_0$), because if $r_{ave} = p_1 r_1 + p_0 r_0 = p_1 r_0 RR + p_0 r_0$,we can estimate $r_0 = r_{ave} / (p_1 RR + p_0)$

This generalises to categorical risk factors with more than two categories, in this case, the area socioeconomic indicator. Given

$r_{ave} = r_1 \sum_{i=1}^G p_i RR_i$

we can compute $r_1$ given $r_{ave}$, the socioeconomic relative rates $RR_i$ and socioeconomic population proportions $p_i$. Hence we can compute the remaining $r_i$ by multiplying $r_1$ by $RR_i$.

## Implementation of disaggregating risks for cancer incidence

We assume that $p_i = 1/5$ is the proportion of people in each of the five quintiles of socioeconomic status, and these proportions are the same for each age/sex group. The quintiles are defined so that the same number of areas falls in each quintile. So while this assumption will not be exact, it is expected to be approximately true. To get better estimates of these proportions, we would need data on the population by age, sex and SES quintile.

The code to perform this disaggregation works by stretching out the age/sex incidence data vertically to create a dataset with one row per combination of age, sex and socioeconomic quintile. This is then joined with the dataset `cancerinc_socio` on relative incidence by socioeconomic quintiles. The term $\sum_{i=1}^G p_i RR_i$ is then computed as `sum_pRR`, and joined to the streched dataset `cancerinc`. This allows the disaggregated rates `rate` to be computed in a vectorised way.

```{r}
cancerinc <- cancerinc_agesex |> 
    slice(rep(row_number(), each=5)) |>
    mutate(socio = rep(1:5, length.out=n())) |>
    left_join(cancerinc_socio |> select(site, socio, rr_socio = rr), 
              by=c("site","socio")) |>
    mutate(p_socio = 1/5,    # assumption
           pRR = p_socio * rr_socio,
           socio = ordered(socio))

sum_p <- cancerinc |> 
    group_by(site, age, sex) |>
    summarise(sum_pRR = sum(pRR), .groups="drop")
    
cancerinc <- cancerinc |>
    left_join(sum_p, by=c("site","age","sex")) |>
    mutate(rate_soc = rate * rr_socio / sum_pRR, # Change to rr_soc to keep original value (BZ-D 13-03-2024)
           socio = ordered(socio, labels = c("1 (most deprived)",
                                             "2","3","4",
                                             "5 (least deprived)")),
           sex = factor(sex),
           sex = fct_recode(sex, Female="Females", Male="Males")) #|>
   # select(site, age, sex, socio, rate) # keeping all variables to track calculations (BZ-D 13-03-02024)
saveRDS(cancerinc, file="cancer_incidence_test.rds")
```

We are left with the incidence by age, sex and socioeconomic status in the variable `rate`. The final dataset is saved in the file `cancer_incidence.rds`, and the first few rows are shown here.

```{r}
knitr::kable(head(cancerinc))
```

# Illustration of incidence estimates

Incidence by age, sex and socioeconomic quintile for all cancers considered in JIBE.

The published incidence by age and sex (averaged over deprivation levels) is shown by the dotted black lines. The estimated incidence disaggregated by socioeconomic quintile is shown by the coloured lines, with socioeconomic status indicated by the lightness of the lines.

This shows that the relative risk between socioeconomic quintiles is assumed to be the same between different age/sex groups, but the average of the deprivation-specific estimates is constrained to be the same as the published average.

Note the varying socioeconomic gradients, e.g. breast cancer is commonest among the least deprived, and lung cancer commonest for the most deprived.

For some cancers (esophageal, gastric, myeloid leukemia, myeloma) the apparent lack of socioeconomic effect is due to the assumption that this effect is the same as for all cancers combined, which may not be true. Perhaps this effect might be borrowed from other countries, if there is data.

```{r}
#| fig-height: 12
ggplot(cancerinc, 
       aes(x=age, y=rate_soc, col=sex, alpha=socio)) + 
    geom_line(lwd=1.5) + 
    geom_line(data=cancerinc_agesex,
              aes(x=age, y=rate, group=sex), 
              inherit.aes = FALSE,
              lwd=1, lty=2, alpha=0.4, col="black") +
    facet_wrap(~site, ncol=2, scales="free_y") + 
    ylab("Incidence rate (per 100,000)") +
    xlab("Age (years)") +
    coord_cartesian(xlim=c(25, 90)) +
    guides(alpha=guide_legend(title=NULL),
           col=guide_legend(title=NULL)) +
    theme(legend.position = "bottom")
```

```         
```

# Alternative data sources

We have data from the Cancer Atlas, by SA2 area and for males, females and persons. This is modelled data and methods are explained here ([https://atlas.cancer.org.au/developing-a-cancer-atlas/](https://aus01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fatlas.cancer.org.au%2Fdeveloping-a-cancer-atlas%2F&data=05%7C02%7Cbelen.zapata-diomedi%40rmit.edu.au%7Ccfb444d6cd46424fd54408dc3e4ec630%7Cd1323671cdbe4417b4d4bdb24b51316b%7C0%7C0%7C638453755273552440%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=6dPCoTvHHLX%2Fg8sCOPRkhAfJ2mg%2FA%2FqhSxPbp6mOpbo%3D&reserved=0)) and here ([https://pubmed.ncbi.nlm.nih.gov/31570101/](https://aus01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fpubmed.ncbi.nlm.nih.gov%2F31570101%2F&data=05%7C02%7Cbelen.zapata-diomedi%40rmit.edu.au%7Ccfb444d6cd46424fd54408dc3e4ec630%7Cd1323671cdbe4417b4d4bdb24b51316b%7C0%7C0%7C638453755273491525%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=i25wOiLsG6QQALarEfdMhjFrmdt%2BT3EEFroXhl5ZRxI%3D&reserved=0)). Data is available for the below cancer types and by SA2 area defined by the Australian Statistical Geography Standard July 2011 edition, with data on Capital Cities for Australia so we can select Melbourne. I joined all diseases data for incidence, for males and females, by SA2 area. Data is for standardised incidence ratio (SIR). The SIR reflects the area-specific incidence rate relative to the Australian average. It is the ratio of the observed cancer cases to the expected number of cases, the latter adjusting for difference in population between SA2 and differences in age structure of the population with an SA2

```{r}

# Get sheet with SA2 areas by SES and filter Victoria


library(readxl)

excel_file <- "health/original/ATLAS/Atlas_estimates_95CIs.xlsx"

# Get the names of the sheets in the Excel workbook
diseases <- excel_sheets(excel_file)[-c(1:4)]

## Males
all_data_males <- list()

# Loop through each disease
for (sheet_name in diseases) {
  # Read data for the current disease
  data <- read_xlsx(excel_file, 
    sheet = sheet_name, 
    skip = 9, 
    col_names = TRUE, 
    n_max = 2200
  ) 
  
  # Process data for males
  males <- data %>%
    select(1:5) %>%
    setNames(.[2, ]) %>%
    slice(-1, -2) %>%
    mutate(sex = "Males",
           site = sheet_name)
  
  # Store the processed data in the list
  all_data_males[[sheet_name]] <- males
}

# Bind all data frames together
final_data_males <- do.call(rbind, all_data_males)



## females
all_data_females <- list()

# Loop through each disease
for (sheet_name in diseases) {
  # Read data for the current disease
  data <- read_xlsx(excel_file, 
    sheet = sheet_name, 
    skip = 9, 
    col_names = TRUE, 
    n_max = 2200
  ) 
  
  # Process data for males
  females <- data %>%
    select(1,2,6,7,8) %>%
  setNames(.[2, ]) %>%
  slice(-1, -2) %>%
  mutate(sex="Females",
           site = sheet_name)
  
  # Store the processed data in the list
  all_data_females[[sheet_name]] <- females
}

# Bind all data frames together
final_data_females <- do.call(rbind, all_data_females) 

SA2_soc <- read_xlsx(excel_file, sheet="Geography", skip=11, col_names = TRUE, n_max=2200)

data_combined <- bind_rows(final_data_females, final_data_males) %>%
  mutate(`SA2 code`=as.numeric(`SA2 code`)) %>%
 left_join(SA2_soc) %>% rename(SA2 = `SA2 name`)
# %>% filter(State=="Victoria", Remoteness=="Major city", `Capital cities`=="Greater Melbourne")
```

# Combining data sources

I combined incidence data by age and sex for Australia wide with data for SA2 area. Data for SA2 area (data_combined) from the Atlas gives us the SIR, which is interpreted as the relative rate of incidence for an area, by sex to the Australian average. The Australian average is from AIHW and already stretched to one year age groups in data inc.

Table to match ICD10 codes names in age and sex data for data file inc.

```{r}
sites <- rbind( 
    ## Standard name,   Name in age/sex data,  Name in ATLAS,   Our intended ICD10,
    c("Breast cancer",   "Breast cancer"  ,      "Breast cancer (C50)" , "C50"),
     c("Endometrial cancer", "Endometrial cancer", "Uterine cancer (C54–C55)" ,  "C54–C55"),
      c("Lung cancer",       "Lung cancer",         "Lung cancer (C33–C34)",  "C33–C34"),
      c("Colon cancer",     "Colon cancer",         "Bowel cancer (C18–C20)", "C18"),
      c("Esophageal cancer", "Oesophageal cancer" , "Oesophageal cancer (C15)",  "C15"),  
      c("Liver cancer",      "Liver cancer",       "Liver cancer (C22)", "C22"),
      c("Kidney cancer",     "Kidney cancer",    "Kidney cancer (C64)", "C64"),
      c("Gastric cardia cancer",  "Stomach cancer", "Stomach cancer (C16)", "C16"),
      c("Myeloid leukemia", "Chronic myeloid leukaemia (CML)"  , "Leukaemia (C91-C95)", "C92.1"),
      c("Myeloma", "Multiple myeloma", "Myeloma (C90)", "C90.0"),
      c("Head and neck cancer", "Head and neck cancer (including lip)", "Head and neck cancers (with lip) (C00–C14, C30–C32)", "C01–C14, C30–C32"),
      c("Rectum cancer", "Rectal cancer (including rectosigmoid junction)", "Bowel cancer (C18–C20)", "C20"),
      c("Bladder cancer", "Bladder cancer", NA, "C67"),
      c("Prostate cancer", "Prostate cancer", "Prostate cancer (C61)", "C61"),
      c("Malignant melanoma", "Melanoma of the skin", "Melanoma of the skin (C43)", "C43"),
    c("All cancers", "All cancers combined", "All cancers combined (C00–C97, D45, D46, D47.1, D47.3–D47.5)", "C00–C97, D45, D46, D47.1, D47.3–D47.5")
    ) |> 
    as_tibble() |>
    setNames(c("stdname","agesexname","socioname","ICD10") ) |>
    select(stdname, ICD10, agesexname, socioname)


# Generate the table using knitr::kable
knitr::kable(sites, col.names = c("Standardised name","ICD 10 code",
             "Closest site in age/sex data","Closest site in socioeconomic data"))

```

Select data for health model in JIBE. Rate by sex, age and area is simply created by multiplying Standardized Incidence Rate (SIR) by age and sex rate.

```{r}

#Select diseases from ATLAS combined data
data_combined <- data_combined %>% filter(!(site %in% c("Brain Cancer", "Cervical Cancer", "Kidney Cancer", "Myeloproliferative neoplasms", "Non-Hodgkin Lymphoma", "Ovarian Cancer", "Pancreatic Cancer", "Thyroid Cancer"))) %>% mutate(site=tolower(site)) 
# 

# Atlas does not have data for bladder cancer, use data for all cancers and for colon and rectum separetly, then us combined (bowel cancer) for each. 

bladder_cancer <- data_combined %>% filter(site=="all cancers") %>%
  mutate(site = if_else(site == "all cancers", "bladder cancer", site))

rectum_cancer <- data_combined %>% filter(site=="bowel cancer") %>%
  mutate(site = if_else(site == "bowel cancer", "rectum cancer", site))
  
data_sa2_sex <- data_combined %>%
  mutate(site = if_else(site == "bowel cancer", "colon cancer", site)) %>%
           bind_rows(bladder_cancer, rectum_cancer) %>%
  mutate(site = if_else(site == "uterine cancer", "endometrial cancer", site),
         site = if_else(site == "oesophageal cancer", "esophageal cancer", site),
         site = if_else(site == "melanoma", "malignant melanoma", site),
         site = if_else(site == "stomach cancer", "gastric cardia cancer", site), 
         site = if_else(site == "leukaemia", "myeloid leukemia", site),
         site = ifelse(site == "head and neck cancers", "head and neck cancer", site)) 


## Combine with incidence data by age and sex

inc <- inc %>% mutate(site=tolower(site))

data_sa2_sex_age <- data_sa2_sex %>% left_join(inc) %>%
  mutate(rate_area=as.numeric(SIR)*rate) %>%
  rename(socio=`Socioeconomic status`)

## Compare data per SA2 and sex and data from national administrative sources by age, sex and combined with level of deprivation (Chris)



compare_socio <- data_sa2_sex_age %>%
group_by(age, sex, site, socio, .groups = "drop") %>%
summarise(rate_area = round(mean(rate_area, na.rm = TRUE), 1), .groups = "drop") %>% mutate(socio=as.factor(socio)) %>%
  filter(socio != "Not classified") %>%
  mutate(sex = case_when(
    sex == "Females" ~ "Female",
    sex == "Males" ~ "Male"
  ))

cancerinc <- cancerinc %>%
  mutate(site = tolower(site)) %>%
  mutate(socio = case_when(
    socio == "1 (most deprived)" ~ "Most disadvantaged",
    socio == "2" ~ "Disadvantaged",
    socio == "3" ~ "Middle SES", 
    socio == "4" ~ "Advantaged",
    socio == "5 (least deprived)" ~ "Most advantaged",
    TRUE ~ socio  # Handle default case
  )) %>%
mutate(socio=as.factor(socio))

compare_socio <- compare_socio %>% left_join(cancerinc) %>%
  mutate(rate_soc=round(rate_soc, 1))
```

Each graph created below for each site and sex combinations depicts five graphs within representing each socioeconomic group, and within each graph by socioeconomic group a line called rate which represents the rate by age and sex for Australia wide (inc), the rate derived from age and sex and age standardised rates by socioeconomic group called rate_soc (derived by Chris) and rate derived from data by sex and SA2 and rates by age and sex (rate_area).

```{r}


rates_compare_long <- compare_socio %>% pivot_longer(
  cols = c(rate_area, rate, rate_soc),
  values_to = "rate",
  names_to = "rate_type"
)

library(ggplot2)

# Create folder if it doesn't exist
folder_path <- file.path("docs", "images")
if (!file.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
}

# Define colors for rate types
rate_type_colors <- c("black", "red", "blue")

# Loop over each combination of site and sex
for (site_val in unique(rates_compare_long$site)) {
  for (sex_val in unique(rates_compare_long$sex)) {
    # Skip certain combinations of sex and site
    if ((sex_val == "Male" && site_val %in% c("breast cancer", "endometrial cancer")) ||
        (sex_val == "Female" && site_val == "prostate cancer")) {
      cat("Skipping graph for sex:", sex_val, "and site:", site_val, "\n")
      next  # Skip to the next iteration of the loop
    }
    
    cat("Processing site:", site_val, "and sex:", sex_val, "\n")  # Debug print statement
    
    # Subset data for the current combination of site and sex
    plot_data <- subset(rates_compare_long, site == site_val & sex == sex_val)
    
    # Check if there are data for the current combination
    if (nrow(plot_data) > 0) {
      max_rate <- max(plot_data$rate, na.rm = TRUE)
      cat("Maximum rate:", max_rate, "\n")  # Debug print statement
      
      # Create ggplot for the current combination
      p <- ggplot(plot_data, aes(x = age, y = rate, col = rate_type)) +
        geom_line(lwd = 1, aes(linetype = NULL)) +
        scale_color_manual(values = rate_type_colors) +
        facet_wrap(~socio, scales = "free_y", ncol = 1) + # Nested facet wrap for socio-economic status
        ylab("Incidence rate (per 100,000)") +
        xlab("Age (years)") +
        coord_cartesian(xlim = c(25, 90)) +
        scale_y_continuous(breaks = seq(0, max_rate, by = 500)) +  # Adjust y-axis breaks
        guides(alpha = FALSE, col = guide_legend(title = "Rate Type")) +
        theme(legend.position = "bottom", legend.box = "horizontal", legend.box.just = "center",
              legend.margin = margin(t = 5, unit = "pt"), legend.spacing = unit(0.2, "cm"))
      
      # Save the ggplot for the current combination
      png_name <- file.path(folder_path, paste0(site_val, "_", sex_val, ".png"))
      ggsave(png_name, plot = p + labs(title = paste("| Site:", site_val, "| Sex:", sex_val)), device = "png")
    } else {
      cat("No data for site:", site_val, "and sex:", sex_val, "\n")  # Debug print statement
    }
  }
}


# Second comparison difference between rate_soc and rate area

compare_diff <- compare_socio %>% 
  mutate(diff_rate = ifelse(rate_area == 0 | rate_soc == 0, 0, ((rate_area - rate_soc) / rate_soc) * 100)) 
```

# Notes on comparison data sources

In general, rates calculated by Chris with AIHW data and rates derived from ATLAS data are very similar, with small discrepancies for cancer with small incidence rates.

Esophageal cancer rates for males are higher and lower respectively for disadvantaged and most disadvantage and most advantaged when usisng rates based on areas (ATLAS) data. Rates from socioeconomic status (from AIHW data) for esophageal cancer were assumed to follow the same patterns as all cancers.

For gastric cardia, for both males and females, rates based on areas (ATLAS) are slightly lower than from socio-economic data (AIHW) for males and females disadvantaged, most advantaged and advantaged groups.

For head and neck, rates for males and females from area rates (ATLAS) are more conservative than with AIHW, for example, for the advantaged group, area rates are higher than rates from socioeconomic data (AIHW). Similar pattern with liver cancer and lung cancer.
