---
title: "JIBE Melbourne: Cardiovascular incidence data"
author: "Chris Jackson, Belen Zapata-Diomedi"
email: "chris.jackson@mrc-bsu.cam.ac.uk"
date: today
format: 
  html:
    code-fold: true
    embed-resources: true
    df-print: paged
execute: 
    message: false
    warning: false
toc: true
number-sections: true
editor: visual
---

This script produces estimates of stroke incidence, ischemic heart disease and type 2 diabetes by age, sex and area socioeconomic quintile, for Australia using data for 2016 and 2018.

The data is obtained from the Global Burden of Disease for disease incidence by age and sex and from AIHW by sex and socioeconomic status ([JIBE data summary spreadsheet](https://gbr01.safelinks.protection.outlook.com/ap/x-59584e83/?url=https%3A%2F%2Frmiteduau.sharepoint.com%2F%3Ax%3A%2Fr%2Fsites%2FJIBEUKAUS%2FShared%2520Documents%2FWP%25206%2Fhealth%2Fdata%2Fdiseases%2520data.xlsx%3Fd%3Dwed9fbef6faa046df8f4e6453f1a1463b%26csf%3D1%26web%3D1%26e%3DhKUlh4&data=05%7C02%7Cchris.jackson%40mrc-bsu.cam.ac.uk%7Cdfd4effdc1c346447c0508dc30f7a9ae%7C513def5bdf174107b5523dba009e5990%7C0%7C0%7C638439087576234119%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=WdIPqRrgjJze3vSbAhvWQYZO3K1R2IUlDXcxbWXFVJA%3D&reserved=0)). There is no unique source of data for what we need, hence the combination of data sources. Source files:

-   Incidence by 5-year age group and sex for stroke, ischemic heart disease and type 2 diabetes:

    -   gbd_incidence.csv (available from GBD data tool, data is for 2018, baseline year for the jibe model)

-   Incidence, death and prevalence by sex and socioeconomic area for 2016. We do not have data for incidence by socioeconomic area for all diseases of interest, hence, we use deaths and prevalence, depending which one is more suitable.

    -   cvd_dia_kid_socioeconomic.xlsx, which contains:

1.  Stroke incidence: by sex and socioeconomic area for 2016: Table S2

2.  Cardiovascular disease death: by sex and socioeconomic area for 2016: Table S2

3.  Type 2 diabetes prevalence: by sex and socioeconomic area for 2016: Table S7

These data refer to the whole of Australia, and we do not have area specific data (e.g. SA2 as we do with mortality). These data are combined to produce estimates for each of stroke, ischemic heart disease and type 2 diabetes by age, sex and area socioeconomic quintile. This assumes that the effect of socioeconomic status on stroke, ischemic heart disease and diabetes type 2 incidence is the same for all age and sex groups.

```{r}
library(tidyverse)
library(readxl)
```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/beluz/OneDrive - RMIT University/JIBE/JIBE-WP6/healthmicrosim/")
knitr::opts_chunk$set(cache = TRUE)
```

# Incidence by socioeconomic status and sex

-   Socioeconomic areas are classified according to population-based quintiles using the Index of Relative Socio-Economic Disadvantage (IRSD) based on Statistical Area Level 2 (SA2) of usual residence
-   The below data sources are then used to adjust GBD data by age and sex to also be further dissagregated by socioeconomic status.

```{r}
# Stroke

stroke_inc <- read_xlsx("health/original/AIHW/cdv_dia_kid_socioeconomic.xlsx", sheet="Table S2", skip=1, col_names = TRUE, n_max=30) %>%
    slice(-c(7:11, 18:26)) %>%
    mutate(sex = ifelse(row_number() <= 6, "Male", "Female")) %>%
    slice(-c(1,7)) %>%
    rename(socio=`Socioeconomic group`, 
           rate=`2016`) %>%
    mutate(socio = case_when(socio=="Group 1 (lowest)" ~ "1", # 1: lowest, to 5: highest
                             socio=="Group 2" ~ "2",
                             socio=="Group 3" ~ "3",
                             socio=="Group 4" ~ "4",
                             socio=="Group 5 (highest)" ~ "5",
                             .default = socio),
           socio = as.numeric(socio)) %>%
  group_by(sex) %>%
  mutate(rr = rate / first(rate)) |>
  ungroup() %>%
  select(socio, sex, rate, rr) %>%
  mutate(cause="Stroke") # to match gbd name

# Cardiovascular disease (data is for deaths but we will use it to adjust incidence by age and sex)

cvd_inc <- read_xlsx("health/original/AIHW/cdv_dia_kid_socioeconomic.xlsx", sheet="Table S3", skip=1, col_names = TRUE, n_max=30) %>%
    slice(-c(7:11, 18:26)) %>%
    mutate(sex = ifelse(row_number() <= 6, "Male", "Female")) %>%
    slice(-c(1,7)) %>%
    rename(socio=`Socioeconomic group`, 
           rate=`2016`) %>%
    mutate(socio = case_when(socio=="Group 1 (lowest)" ~ "1", # 1: lowest, to 5: highest
                             socio=="Group 2" ~ "2",
                             socio=="Group 3" ~ "3",
                             socio=="Group 4" ~ "4",
                             socio=="Group 5 (highest)" ~ "5",
                             .default = socio),
           socio = as.numeric(socio)) %>%
  group_by(sex) %>%
  mutate(rr = rate / first(rate)) |>
  ungroup() %>%
  select(socio, sex, rate, rr) %>%
  mutate(cause="Ischemic heart disease") # to match gbd name


# Diabetes (we use prevalence data. There is mortlaity data, but figures are low and diabetes as cause of deaths is likely under reported)

diabetes_inc <- read_xlsx("health/original/AIHW/cdv_dia_kid_socioeconomic.xlsx", sheet="Table S7", skip=1, col_names = TRUE, n_max=30) %>%
    slice(-c(7:11, 18:26)) %>%
    mutate(sex = ifelse(row_number() <= 6, "Male", "Female")) %>%
    slice(-c(1,7,13)) %>%
    rename(socio=`Socioeconomic group`, 
           rate=`2016`) %>%
    mutate(socio = case_when(socio=="Group 1 (lowest)" ~ "1", # 1: lowest, to 5: highest
                             socio=="Group 2" ~ "2",
                             socio=="Group 3" ~ "3",
                             socio=="Group 4" ~ "4",
                             socio=="Group 5 (highest)" ~ "5",
                             .default = socio),
           socio = as.numeric(socio)) %>%
  group_by(sex) %>%
  mutate(rr = rate / first(rate)) |>
  ungroup() %>%
  select(socio, sex, rate, rr) %>%
  mutate(cause="Diabetes mellitus type 2") #to match gbd name

# Combined database with all three diseases

cvd_dia_socio <- bind_rows(cvd_inc, diabetes_inc, stroke_inc)
```

# Estimating incidence by combinations of risk factors

As in the Melbourne mortality data analysis (section 2.2) to disaggregate an event rate (incidence by age and sex in this case) by a further risk factor, given

-   the rate $r_{ave}$ by age and sex, averaged over people with or without the risk factor

-   a relative (unstandardised) rate $RR$ between people with/without the risk factor

-   the proportions of people with the risk factor $p_1$ and without the risk factor $p_0$

we can estimate the rates with ($r_1$) and without the risk factor ($r_0$), because if $r_{ave} = p_1 r_1 + p_0 r_0 = p_1 r_0 RR + p_0 r_0$,we can estimate $r_0 = r_{ave} / (p_1 RR + p_0)$

This generalises to categorical risk factors with more than two categories, in this case, the area socioeconomic indicator. Given

$r_{ave} = r_1 \sum_{i=1}^G p_i RR_i$

we can compute $r_1$ given $r_{ave}$, the socioeconomic relative rates $RR_i$ and socioeconomic population proportions $p_i$. Hence we can compute the remaining $r_i$ by multiplying $r_1$ by $RR_i$.

## Implementation of disaggregating risks for stroke, ischemic heart disease and type 2 diabetes incidence

We assume that $p_i = 1/5$ is the proportion of people in each of the five quintiles of socioeconomic status, and these proportions are the same for each age/sex group. The quintiles are defined so that the same number of areas falls in each quintile. So while this assumption will not be exact, it is expected to be approximately true. To get better estimates of these proportions, we would need data on the population by age, sex and SES quintile.

The code to perform this disaggregation works by stretching out the age/sex incidence data vertically to create a dataset with one row per combination of age, sex and socioeconomic quintile. This is then joined with the dataset `cvd_dia_socio` on relative incidence by socioeconomic quintiles. The term $\sum_{i=1}^G p_i RR_i$ is then computed as `sum_pRR`, and joined to the stretched dataset `gbb_inc`. This allows the disaggregated rates `rate` to be computed in a vectorised way.

```{r}

gbd <- read.csv("health/original/GBD/gbd_incidence.csv") %>%
  filter(measure %in% "Incidence") %>%
  filter (cause %in% c("Stroke", "Ischemic heart disease", "Diabetes mellitus type 2")) %>%
  filter(metric %in% "Rate") %>%
  filter(year %in% 2018)

# Strech out the data repeating for each age and sex group by five and then generate additional columns indication proportion of people in socioeconomic category (p_socio), pRR (prevalence by the RR for fcalculation of sum pRR as explained above)

gbd_inc <- gbd |> 
    slice(rep(row_number(), each=5)) |>
    mutate(socio = rep(1:5, length.out=n())) |>
    left_join(cvd_dia_socio |> select(cause, socio, sex, rr_socio = rr), 
              by=c("cause","socio", "sex")) |>
    mutate(p_socio = 1/5,    # assumption
           pRR = p_socio * rr_socio,
           socio = ordered(socio))


# Calculate sum_pRR

sum_p <- gbd_inc |> 
    group_by(cause, age, sex) %>%
    summarise(sum_pRR = sum(pRR), .groups="drop")
    
gbd_inc_soc <- gbd_inc %>%
  left_join(sum_p, by = c("cause", "age", "sex")) %>%
  mutate(rate_soc = val * rr_socio / sum_pRR,
         socio = ordered(socio, labels = c("1 (most deprived)",
                                           "2", "3", "4",
                                           "5 (least deprived)")),
         sex = factor(sex),
         age = if_else(age == "95+ years", "95-99",
                       if_else(age == "<5 years", "0-4", age)),
         age = str_replace(as.character(age), " years", ""),
         rate=rate_soc/100000, 
         prob=1 - exp(-rate)) 




## Dissagregate in 1 year age groups, for modelling we might prefer 1 year age group, but check with Ali, might be easier for the microsimulation to pick from an age range for an individual otherwise lot of data. Here we create both, single year and 5 year age group. 

### Simple method repeat age group value for each one year age group

## Working with the data that has one row per five-year age group, 
## first construct 1-year counts as an extra column. 

gbdp <- gbd_inc_soc %>%
    tidyr::separate(age, into = c("from_age", "to_age"), sep = "-", convert = TRUE, remove = FALSE) %>%
    mutate(agediff = as.numeric(to_age) - as.numeric(from_age) + 1) %>%
    rename(agegroup=age)


## Now stretch the data out using an index, to create a data frame with 1 row per year of age and create a variable for year of age. 

index <- rep(1:nrow(gbdp), gbdp$agediff)
gbdpyrd5 <- gbdp[index,] %>%
    mutate(age = from_age + sequence(gbdp$agediff) - 1)

## Now this is the one year data set with simple expansion for age groups. 
gbdpyrd5 <- gbdpyrd5 


# #### To do later 
# ## More advanced method for smooth disaggregation 
# ## using the tempdisagg package 
# # tmp <- gbdp %>% group_by(measure, location, sex, cause) %>% filter(cur_group_id()==1)
# 
# gbdp_grp <- gbdp %>% 
#   group_by(measure, sex, cause) %>%  # filter(cur_group_id() %in% 1:3) %>%
#   arrange(measure, sex, cause, from_age)
# 
# disagg <- function(dat, key){ 
#   res <- with(dat, { 
#     data.frame( 
#       ageyr = rep(from_age,agediff) + sequence(agediff) - 1,
#       val = predict(td(val ~ 1, to=5, method="fast"))
#     ) } )
#   res 
# }
# 
# # This takes about a minute 
# library(tempdisagg)
# gbdpyr <- group_modify(gbdp_grp, disagg) %>% 
#   ungroup %>%
#   left_join(gbdpyrd5, by=c("measure","ageyr","sex","cause"))
# 
# # Sometimes the results are negative.  
# # Revert to dividing by 5 for all agegroups that contain a negative result 
# # neg_val <- gbdpyr %>%
# #   filter(val<0) %>%
# #   select("measure","sex","cause","agegroup") %>%
# #   distinct() %>%
# #   mutate(zeronum = TRUE)
# # 
# # gbdpyr <- gbdpyr %>% 
# #   left_join(neg_val, by=c("measure","sex","cause","agegroup")) %>%
# #   mutate(zeronum = replace_na(zeronum, FALSE)) %>%
# #   mutate(zerodenom = replace_na(zerodenom, FALSE)) %>%
# #   mutate(val= if_else(zeronum, num1yr, round(num)),
# #          denom = if_else(zerodenom, denom1yr, round(denom)))

saveRDS(gbdpyrd5, file="health/processed/cvd_dia_incidence_socio.rds")
```

We are left with the incidence by age, sex and socioeconomic status in the variable `rate`. The final dataset is saved in the file `cvd_dia_incidence_socio.rds`, and the first few rows are shown here.

```{r}
knitr::kable(head(gbdpyrd5))
```

# Illustration of incidence estimates

Incidence by age, sex and socioeconomic quintile for ischemic heart disease, stroke and type 2 diabetes considered in JIBE.

The published incidence by age and sex (averaged over deprivation levels) is shown by the dotted black lines (original gbd data streched from 5 year age groups to one year). The estimated incidence disaggregated by socioeconomic quintile is shown by the coloured lines, with socioeconomic status indicated by the lightness of the lines (calculated from gbd data and aihw data).

This shows that the relative risk between socioeconomic quintiles is assumed to be the same between different age/sex groups, but the average of the deprivation-specific estimates is constrained to be the same as the published average.

Comments need updating.

**Note the varying socioeconomic gradients, e.g. breast cancer is commonest among the least deprived, and lung cancer commonest for the most deprived.**

**For some cancers (esophageal, gastric, myeloid leukemia, myeloma) the apparent lack of socioeconomic effect is due to the assumption that this effect is the same as for all cancers combined, which may not be true. Perhaps this effect might be borrowed from other countries, if there is data.**

```{r}
#| fig-height: 12

gbdpyrd5_figure <- gbdpyrd5 %>%
  mutate(rate=rate*100000)

# Load necessary libraries
library(dplyr)
library(tidyr)
library(stringr)

# Step 1: Transform the age variable and calculate agediff
gbd_inc_expanded <- gbd_inc %>%
  mutate(age = case_when(
    age == "<5 years" ~ "0-4",
    age == "95+ years" ~ "95-99",
    TRUE ~ age
  )) %>%
  mutate(age = str_replace(age, " years", "")) %>%  # Remove the word "years"
  separate(age, into = c("from_age", "to_age"), sep = "-", convert = TRUE, remove = FALSE) %>%
  mutate(agediff = as.numeric(to_age) - as.numeric(from_age) + 1) %>%
  rename(agegroup = age)

# Step 2: Stretch the data out using an index
index <- rep(1:nrow(gbd_inc_expanded), gbd_inc_expanded$agediff)
gbd_inc_expanded <- gbd_inc_expanded[index,]

# Step 3: Create a variable for the year of age
gbd_inc_expanded <- gbd_inc_expanded %>%
  group_by(row_number()) %>%
  mutate(age = from_age + sequence(n()) - 1) %>%
  ungroup() %>%
  rename(rate=val)

# Debugging: Check the final dataframe
print(head(gbd_inc_expanded))



## Prepare plot

library(ggplot2)

# Plot
ggplot(gbdpyrd5_figure, 
       aes(x = age, y = rate, col = sex, alpha = socio)) + 
    geom_line(lwd = 1.5) +  # Solid lines for gbdpyrd5_figure
    geom_line(data = gbd_inc_expanded,
              aes(x = age, y = rate, col = sex, group = sex),  # Use col and group aesthetics
              inherit.aes = FALSE,
              lwd = 1, lty = 2, alpha = 0.4) +  # Dotted lines for gbd_inc_expanded
    facet_wrap(~cause, ncol = 1, scales = "free_y") +
    ylab("Incidence rate (per 100,000)") +
    xlab("Age (years)") +
    coord_cartesian(xlim = c(25, 90)) +
    guides(alpha = guide_legend(title = NULL),
           col = guide_legend(title = NULL)) +
    theme(legend.position = "bottom")

```

```         
```

# Alternative data sources

We have data from the Cancer Atlas, by SA2 area and for males, females and persons. This is modelled data and methods are explained here ([https://atlas.cancer.org.au/developing-a-cancer-atlas/](https://aus01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fatlas.cancer.org.au%2Fdeveloping-a-cancer-atlas%2F&data=05%7C02%7Cbelen.zapata-diomedi%40rmit.edu.au%7Ccfb444d6cd46424fd54408dc3e4ec630%7Cd1323671cdbe4417b4d4bdb24b51316b%7C0%7C0%7C638453755273552440%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=6dPCoTvHHLX%2Fg8sCOPRkhAfJ2mg%2FA%2FqhSxPbp6mOpbo%3D&reserved=0)) and here ([https://pubmed.ncbi.nlm.nih.gov/31570101/](https://aus01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fpubmed.ncbi.nlm.nih.gov%2F31570101%2F&data=05%7C02%7Cbelen.zapata-diomedi%40rmit.edu.au%7Ccfb444d6cd46424fd54408dc3e4ec630%7Cd1323671cdbe4417b4d4bdb24b51316b%7C0%7C0%7C638453755273491525%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=i25wOiLsG6QQALarEfdMhjFrmdt%2BT3EEFroXhl5ZRxI%3D&reserved=0)). Data is available for the below cancer types and by SA2 area defined by the Australian Statistical Geography Standard July 2011 edition, with data on Capital Cities for Australia so we can select Melbourne. I joined all diseases data for incidence, for males and females, by SA2 area. Data is for standardised incidence ratio (SIR). The SIR reflects the area-specific incidence rate relative to the Australian average. It is the ratio of the observed cancer cases to the expected number of cases, the latter adjusting for difference in population between SA2 and differences in age structure of the population with an SA2

```{r}

# Get sheet with SA2 areas by SES and filter Victoria


library(readxl)

excel_file <- "health/original/ATLAS/Atlas_estimates_95CIs.xlsx"

# Get the names of the sheets in the Excel workbook
diseases <- excel_sheets(excel_file)[-c(1:4)]

## Males
all_data_males <- list()

# Loop through each disease
for (sheet_name in diseases) {
  # Read data for the current disease
  data <- read_xlsx(excel_file, 
    sheet = sheet_name, 
    skip = 9, 
    col_names = TRUE, 
    n_max = 2200
  ) 
  
  # Process data for males
  males <- data %>%
    select(1:5) %>%
    setNames(.[2, ]) %>%
    slice(-1, -2) %>%
    mutate(sex = "Males",
           site = sheet_name)
  
  # Store the processed data in the list
  all_data_males[[sheet_name]] <- males
}

# Bind all data frames together
final_data_males <- do.call(rbind, all_data_males)



## females
all_data_females <- list()

# Loop through each disease
for (sheet_name in diseases) {
  # Read data for the current disease
  data <- read_xlsx(excel_file, 
    sheet = sheet_name, 
    skip = 9, 
    col_names = TRUE, 
    n_max = 2200
  ) 
  
  # Process data for males
  females <- data %>%
    select(1,2,6,7,8) %>%
  setNames(.[2, ]) %>%
  slice(-1, -2) %>%
  mutate(sex="Females",
           site = sheet_name)
  
  # Store the processed data in the list
  all_data_females[[sheet_name]] <- females
}

# Bind all data frames together
final_data_females <- do.call(rbind, all_data_females) 

SA2_soc <- read_xlsx(excel_file, sheet="Geography", skip=11, col_names = TRUE, n_max=2200)

data_combined <- bind_rows(final_data_females, final_data_males) %>%
  mutate(`SA2 code`=as.numeric(`SA2 code`)) %>%
 left_join(SA2_soc) %>% rename(SA2 = `SA2 name`)
# %>% filter(State=="Victoria", Remoteness=="Major city", `Capital cities`=="Greater Melbourne")
```

# Combining data sources

I combined incidence data by age and sex for Australia wide with data for SA2 area. Data for SA2 area (data_combined) from the Atlas gives us the SIR, which is interpreted as the relative rate of incidence for an area, by sex to the Australian average. The Australian average is from AIHW and already stretched to one year age groups in data inc.

Table to match ICD10 codes names in age and sex data for data file inc.

```{r}
sites <- rbind( 
    ## Standard name,   Name in age/sex data,  Name in ATLAS,   Our intended ICD10,
    c("Breast cancer",   "Breast cancer"  ,      "Breast cancer (C50)" , "C50"),
     c("Endometrial cancer", "Endometrial cancer", "Uterine cancer (C54–C55)" ,  "C54–C55"),
      c("Lung cancer",       "Lung cancer",         "Lung cancer (C33–C34)",  "C33–C34"),
      c("Colon cancer",     "Colon cancer",         "Bowel cancer (C18–C20)", "C18"),
      c("Esophageal cancer", "Oesophageal cancer" , "Oesophageal cancer (C15)",  "C15"),  
      c("Liver cancer",      "Liver cancer",       "Liver cancer (C22)", "C22"),
      c("Kidney cancer",     "Kidney cancer",    "Kidney cancer (C64)", "C64"),
      c("Gastric cardia cancer",  "Stomach cancer", "Stomach cancer (C16)", "C16"),
      c("Myeloid leukemia", "Chronic myeloid leukaemia (CML)"  , "Leukaemia (C91-C95)", "C92.1"),
      c("Myeloma", "Multiple myeloma", "Myeloma (C90)", "C90.0"),
      c("Head and neck cancer", "Head and neck cancer (including lip)", "Head and neck cancers (with lip) (C00–C14, C30–C32)", "C01–C14, C30–C32"),
      c("Rectum cancer", "Rectal cancer (including rectosigmoid junction)", "Bowel cancer (C18–C20)", "C20"),
      c("Bladder cancer", "Bladder cancer", NA, "C67"),
      c("Prostate cancer", "Prostate cancer", "Prostate cancer (C61)", "C61"),
      c("Malignant melanoma", "Melanoma of the skin", "Melanoma of the skin (C43)", "C43"),
    c("All cancers", "All cancers combined", "All cancers combined (C00–C97, D45, D46, D47.1, D47.3–D47.5)", "C00–C97, D45, D46, D47.1, D47.3–D47.5")
    ) |> 
    as_tibble() |>
    setNames(c("stdname","agesexname","socioname","ICD10") ) |>
    select(stdname, ICD10, agesexname, socioname)


# Generate the table using knitr::kable
knitr::kable(sites, col.names = c("Standardised name","ICD 10 code",
             "Closest site in age/sex data","Closest site in socioeconomic data"))

```

Select data for health model in JIBE. Rate by sex, age and area is simply created by multiplying Standardized Incidence Rate (SIR) by age and sex rate.

```{r}

#Select diseases from ATLAS combined data
data_combined <- data_combined %>% filter(!(site %in% c("Brain Cancer", "Cervical Cancer", "Kidney Cancer", "Myeloproliferative neoplasms", "Non-Hodgkin Lymphoma", "Ovarian Cancer", "Pancreatic Cancer", "Thyroid Cancer"))) %>% mutate(site=tolower(site)) 
# 

# Atlas does not have data for bladder cancer, use data for all cancers and for colon and rectum separetly, then us combined (bowel cancer) for each. 

bladder_cancer <- data_combined %>% filter(site=="all cancers") %>%
  mutate(site = if_else(site == "all cancers", "bladder cancer", site))

rectum_cancer <- data_combined %>% filter(site=="bowel cancer") %>%
  mutate(site = if_else(site == "bowel cancer", "rectum cancer", site))
  
data_sa2_sex <- data_combined %>%
  mutate(site = if_else(site == "bowel cancer", "colon cancer", site)) %>%
           bind_rows(bladder_cancer, rectum_cancer) %>%
  mutate(site = if_else(site == "uterine cancer", "endometrial cancer", site),
         site = if_else(site == "oesophageal cancer", "esophageal cancer", site),
         site = if_else(site == "melanoma", "malignant melanoma", site),
         site = if_else(site == "stomach cancer", "gastric cardia cancer", site), 
         site = if_else(site == "leukaemia", "myeloid leukemia", site),
         site = ifelse(site == "head and neck cancers", "head and neck cancer", site)) 


## Combine with incidence data by age and sex

inc <- inc %>% mutate(site=tolower(site))

data_sa2_sex_age <- data_sa2_sex %>% left_join(inc) %>%
  mutate(rate_area=as.numeric(SIR)*rate) %>%
  rename(socio=`Socioeconomic status`)

# Save in final folder

cancer_incidence_area <- data_sa2_sex_age %>% rename (sa2_code=`SA2 code`, 
                                                      sa2_name = SA2) %>%
  mutate(prob=1 - exp(-rate_area/100000)) %>%
  select(sa2_code, sa2_name, socio, site, age, sex, prob)

saveRDS(cancer_incidence_area, file="health/processed/cancer_incidence_area.rds")

## Compare data per SA2 and sex and data from national administrative sources by age, sex and combined with level of deprivation (Chris)



compare_socio <- data_sa2_sex_age %>%
group_by(age, sex, site, socio, .groups = "drop") %>%
summarise(rate_area = round(mean(rate_area, na.rm = TRUE), 1), .groups = "drop") %>% mutate(socio=as.factor(socio)) %>%
  filter(socio != "Not classified") %>%
  mutate(sex = case_when(
    sex == "Females" ~ "Female",
    sex == "Males" ~ "Male"
  ))

cancerinc <- cancerinc %>%
  mutate(site = tolower(site)) %>%
  mutate(socio = case_when(
    socio == "1 (most deprived)" ~ "Most disadvantaged",
    socio == "2" ~ "Disadvantaged",
    socio == "3" ~ "Middle SES", 
    socio == "4" ~ "Advantaged",
    socio == "5 (least deprived)" ~ "Most advantaged",
    TRUE ~ socio  # Handle default case
  )) %>%
mutate(socio=as.factor(socio))

compare_socio <- compare_socio %>% left_join(cancerinc) %>%
  mutate(rate_soc=round(rate_soc, 1))
```

Each graph created below for each site and sex combinations depicts five graphs within representing each socioeconomic group, and within each graph by socioeconomic group a line called rate which represents the rate by age and sex for Australia wide (inc), the rate derived from age and sex and age standardised rates by socioeconomic group called rate_soc (derived by Chris) and rate derived from data by sex and SA2 and rates by age and sex (rate_area).

```{r}


rates_compare_long <- compare_socio %>% pivot_longer(
  cols = c(rate_area, rate, rate_soc),
  values_to = "rate",
  names_to = "rate_type"
)

library(ggplot2)

# Create folder if it doesn't exist
folder_path <- file.path("docs", "images")
if (!file.exists(folder_path)) {
  dir.create(folder_path, recursive = TRUE)
}

# Define colors for rate types
rate_type_colors <- c("black", "red", "blue")

# Loop over each combination of site and sex
for (site_val in unique(rates_compare_long$site)) {
  for (sex_val in unique(rates_compare_long$sex)) {
    # Skip certain combinations of sex and site
    if ((sex_val == "Male" && site_val %in% c("breast cancer", "endometrial cancer")) ||
        (sex_val == "Female" && site_val == "prostate cancer")) {
      cat("Skipping graph for sex:", sex_val, "and site:", site_val, "\n")
      next  # Skip to the next iteration of the loop
    }
    
    cat("Processing site:", site_val, "and sex:", sex_val, "\n")  # Debug print statement
    
    # Subset data for the current combination of site and sex
    plot_data <- subset(rates_compare_long, site == site_val & sex == sex_val)
    
    # Check if there are data for the current combination
    if (nrow(plot_data) > 0) {
      max_rate <- max(plot_data$rate, na.rm = TRUE)
      cat("Maximum rate:", max_rate, "\n")  # Debug print statement
      
      # Create ggplot for the current combination
      p <- ggplot(plot_data, aes(x = age, y = rate, col = rate_type)) +
        geom_line(lwd = 1, aes(linetype = NULL)) +
        scale_color_manual(values = rate_type_colors) +
        facet_wrap(~socio, scales = "free_y", ncol = 1) + # Nested facet wrap for socio-economic status
        ylab("Incidence rate (per 100,000)") +
        xlab("Age (years)") +
        coord_cartesian(xlim = c(25, 90)) +
        scale_y_continuous(breaks = seq(0, max_rate, by = 500)) +  # Adjust y-axis breaks
        guides(alpha = FALSE, col = guide_legend(title = "Rate Type")) +
        theme(legend.position = "bottom", legend.box = "horizontal", legend.box.just = "center",
              legend.margin = margin(t = 5, unit = "pt"), legend.spacing = unit(0.2, "cm"))
      
      # Save the ggplot for the current combination
      png_name <- file.path(folder_path, paste0(site_val, "_", sex_val, ".png"))
      ggsave(png_name, plot = p + labs(title = paste("| Site:", site_val, "| Sex:", sex_val)), device = "png")
    } else {
      cat("No data for site:", site_val, "and sex:", sex_val, "\n")  # Debug print statement
    }
  }
}


# Second comparison difference between rate_soc and rate area

compare_diff <- compare_socio %>% 
  mutate(diff_rate = ifelse(rate_area == 0 | rate_soc == 0, 0, ((rate_area - rate_soc) / rate_soc) * 100)) 
```

# Notes on comparison data sources

In general, rates calculated by Chris with AIHW data and rates derived from ATLAS data are very similar, with small discrepancies for cancer with small incidence rates.

Esophageal cancer rates for males are higher and lower respectively for disadvantaged and most disadvantage and most advantaged when usisng rates based on areas (ATLAS) data. Rates from socioeconomic status (from AIHW data) for esophageal cancer were assumed to follow the same patterns as all cancers.

For gastric cardia, for both males and females, rates based on areas (ATLAS) are slightly lower than from socio-economic data (AIHW) for males and females disadvantaged, most advantaged and advantaged groups.

For head and neck, rates for males and females from area rates (ATLAS) are more conservative than with AIHW, for example, for the advantaged group, area rates are higher than rates from socioeconomic data (AIHW). Similar pattern with liver cancer and lung cancer.
